\section{AI for Game Theory}
\label{sec:ai4gt}

\begin{drafting}
\textbf{Part I framing.} AI makes strategic specifications executable. The
recurring move is to relocate a game-theoretic constraint into a differentiable
residual, a no-regret trajectory, an architectural guarantee, or a checkable
certificate. The three subsections take this move in turn: constructive
equilibrium (\ref{subsec:constructive-equilibrium}), strategic inference
(\ref{subsec:strategic-inference}), and neural mechanism design
(\ref{subsec:mechanism-design}).
\cmt[scope]{Selection rule for the whole section: a method enters the main text
only if its objective or stopping condition drives a strategic certificate (Nash
gap / exploitability / regret / IC violation) toward zero, or directly produces
and proves the object. Methods where equilibrium merely emerges as a by-product
(generic return-maximizing self-play or MARL) stay out of the main line.}
\end{drafting}


\subsection{Constructive Equilibrium: Relaxing, Approximating, and Transforming the Nash Certificate}
\label{subsec:constructive-equilibrium}

\begin{drafting}
\emph{Leading question: the Nash equilibrium is easy to define and hard to
compute. Given that the global ``no profitable deviation'' certificate cannot be
enforced directly, what detour around it does each method take?}

\textbf{Thesis.} The difficulty of equilibrium lies in one demand: a global
no-profitable-deviation certificate. Search-space size is not the binding
constraint; the certificate is. Modern AI methods form a spectrum of detours
around it: preserve, residualize, relax in time, substitute the concept,
approximate in space, online the certificate through state transformation, and
re-prove. Across this spectrum, AI (scale, gradients, parameterization) buys
approximation, while the Nash / convergence guarantee always re-enters through a
classical layer (LP, no-regret, safe resolving, symbolic proof).

\begin{itemize}
  \item \textbf{The constructive gap (pre-AI setup).} Existence is free
        (fixed-point theorems); constructing the certificate is the hard,
        regime-dependent step. Finding an exact Nash equilibrium is PPAD-complete
        in general-sum games, while two-player zero-sum collapses to LP /
        no-regret duality~\cite{DaskalakisGoldbergPapadimitriou2009,ChenDengTeng2009}.
        \cmt[layer]{No AI here; this is the complexity partition that governs
        everything downstream.}
  \item \textbf{Transform, no relaxation: classical numerical solvers
        (baseline).} The certificate is rewritten as an algebraic /
        complementarity / LP system (support enumeration, Lemke--Howson,
        homotopy) without weakening it; the price is an explicit, finite, fully
        specified game~\cite{LemkeHowson1964,gambit}.
        \cmt[layer]{Reference zero point: no scale, no gradients, no
        parameterization.}
  \item \textbf{Transform, residualization: differentiable / regularized
        methods.} The certificate becomes the vanishing residual of a
        regularized best-response map, driven to zero by gradients. NashPG fixes
        a large regularization strength and iteratively refines the reference
        policy, yielding monotone improvement and convergence to exact Nash in
        two-player zero-sum without a uniqueness assumption, and scaling to
        domains such as No-Limit Texas Hold'em; QFR gives the first principled
        policy-gradient method with provable best-iterate convergence to a
        regularized Nash in two-player zero-sum imperfect-information EFGs while
        avoiding importance sampling~\cite{NashPG2025,LiuFarinaOzdaglar2025}.
        \cmt[layer]{Purest contemporary AI template: parameterized policy +
        policy gradient + neural networks.}
  \item \textbf{Relax in time: regret minimization.} The one-shot universal
        deviation test relaxes into ``no long-run regret''; exploitability is the
        quantity directly minimized, and equilibrium is the limiting object of a
        learning trajectory. CFR supplies the tabular kernel through
        information-set counterfactual regret; Deep CFR carries the same
        no-regret relaxation into neural function approximation and large
        domains~\cite{Zinkevich2007,BrownDeepCFR2019}.
        \cmt[guardrail]{Restrict to \emph{regret-minimizing} self-play. Pure
        return-maximizing RL self-play (AlphaZero/AlphaStar-type) is excluded:
        equilibrium is only an emergent by-product. The AI lives in the Deep CFR
        approximation layer; the no-regret guarantee comes from learning theory.}
  \item \textbf{Relax the concept: from Nash down to (C)CE.} In general-sum
        games, no-regret dynamics guarantee only a CCE (no-swap-regret a CE):
        polynomially verifiable but weaker certificates. This is a constraint the
        learner inherits, not its contribution. Meta-learned CFR trains a
        parameterized regret minimizer with a strategy-correlation meta-loss that
        steers a CCE-converging learner back toward Nash and bounds the residual
        gap; JPSRO solves equilibrium selection over a deep-RL best-response
        oracle using a max-Gini (C)CE rule~\cite{Marris2021,Sychrovsky2025}.
        \cmt[guardrail]{Say the substitution out loud: JPSRO's ultimate object is
        CCE/CE, not Nash. Verifiability of (C)CE comes from no-regret theory, not
        from deep learning.}
  \item \textbf{Approximate in space: EGTA and PSRO.} When the full game cannot
        be written down, EGTA samples a simulator into an empirical game; PSRO
        then solves a restricted empirical game over a growing policy population,
        using a parameterized deep-RL best-response oracle as the deviation
        search and ``oracle finds no profitable deviation''
        (restricted-game exploitability $\to 0$) as the stopping
        condition~\cite{WellmanTuylsGreenwald2025,McAleerPipelinePSRO2020}. The
        approximation is triple: strategy space, payoffs, and best response.
        \cmt[guardrail]{MARL appears only as PSRO's best-response oracle /
        strategy generator, not as a stand-alone equilibrium-solving direction.}
  \item \textbf{Transform the state and re-prove online: public belief states
        and safe subgame resolving.} An imperfect-information game is lifted to a
        continuous-state perfect-information MDP over public belief states, then
        re-solved online with safe nested subgame resolving, so overall
        exploitability stays under the blueprint; the certificate becomes a
        local, online, error-bounded guarantee. ReBeL learns the value and policy
        networks by self-play deep RL with provable convergence to Nash in
        two-player zero-sum; DeepStack is the deep counterfactual-value
        precursor~\cite{BrownReBeL2020,Moravcik2017}.
        \cmt[guardrail]{Anchor on imperfect-information, exploitability-bounded
        variants. Libratus (CFR+ blueprint + LP-based safe resolving, no neural
        net)~\cite{BrownSandholm2018} appears only as historical contrast;
        perfect-info RL+search is excluded (equilibrium is a by-product).}
  \item \textbf{Re-prove: LLM symbolic solving and algorithm synthesis.}
        Against the trend of lowering the certificate's cost, the LLM branch
        restores it in full, across two layers. Instance solving plus proof:
        PrimeNash derives closed-form Nash and emits machine-checkable proof
        objects~\cite{PrimeNash2025}. Algorithm synthesis: LegoNE compiles a
        Python-like algorithm into a constrained optimization whose solution is a
        provable approximation bound~\cite{LegoNE2025}. AI acts as the engine
        while the certificate is rebuilt symbolically.
        \cmt[guardrail]{Exclude LLM work where equilibrium is merely a tool
        (consensus-game decoding, LLM-as-players evaluation, NE-coordinated
        multi-agent systems).}
\end{itemize}

\textbf{Synthesis.} The detours form a single spectrum: classical
\emph{transform} into an algebraic system $\to$ differentiable \emph{transform}
into a residual fixed point $\to$ \emph{temporal relaxation} into long-run
no-regret $\to$ conceptual \emph{relaxation} into (C)CE $\to$ EGTA/PSRO
\emph{spatial approximation} into an empirical no-deviation population $\to$ PBS
\emph{onlining} of the certificate $\to$ LLM \emph{re-proof}. AI reorganizes the
pipeline through which the Nash certificate is searched, approximated,
substituted, and re-verified, while leaving its rigor intact.
\end{drafting}


\subsection{Strategic Inference: Learning Behavior Generators for Hidden Strategic States}
\label{subsec:strategic-inference}

\begin{drafting}
\emph{Leading question: when observed strategic behavior can be rationalized by
many latent games, what exactly has been identified?}

\textbf{Thesis.} AI turns behavioral theories, game structures, and agent
simulations into trainable \emph{behavior generators} that can be
\emph{constructed}, \emph{inverted}, and \emph{expanded}. Behavioral theory stays
in the loop as the trainable object, so identification keeps its mechanistic
content. The unified problem has two faces: a forward generator
$a_t \sim G_\phi(h_t,z_i)$ and an inversion $p(z_i \mid h_t,a_t)$, where $h_t$ is
observable history and $z_i$ a latent strategic state.
\cmt[axis]{Verb framing that draws the subsection boundaries: 3.2.2 and 3.2.4
are the two ends of the \emph{construct} axis (hand-written theory vs.\ machine
discovery); 3.2.3 in the middle is \emph{invert}. 3.2.4 is motivated by the
limitation 3.2.3 exposes: the true generator may not be in the hand-written
library.}

\begin{itemize}
  \item \textbf{Fit is cheap, identification is hard.} Behavioral fit does not
        amount to mechanism identification. Two underidentification failures must
        be kept apart. \emph{Falsifiability}: does the model restrict behavior at
        all? QRE imposes no falsifiable restriction and can rationalize any
        behavior, so its empirical content comes entirely from auxiliary
        assumptions. \emph{Recovery}: even a structured model may not pin down
        parameters, because the feasible reward set shows a single Nash
        equilibrium mapping to many reward structures, with uniqueness recovered
        only under entropy
        regularization~\cite{HaileHortacsuKosenok2008,FreihautRamponi2025}.
        \cmt[term]{Use ``feasible reward set / reward non-uniqueness under
        equilibrium observation,'' not ``multi-reward underidentification.''}
  \item \textbf{Neuralizing behavioral priors (construct, I).} Theory becomes a
        trainable constraint on the \emph{form} of $G_\phi(h,z)$: a pretraining
        distribution, an architecture, a teacher signal, a regularizer, or a
        candidate family. Cognitive-model priors pretrain on theory-generated
        synthetic data and then fine-tune on human data; GameNet turns
        hand-written cognitive features into a trainable architecture that softly
        encodes iterated-response structure~\cite{Bourgin2019,Hartford2016}.
        \cmt[domain]{Cognitive-model priors are demonstrated on individual risky
        choice (prospect theory), imported here as a paradigm template, not as
        direct game evidence.}
        \cmt[tension]{GameNet self-positions as removing hand-crafted features;
        we read it instead as ``theory as architecture.'' Flag this, and set up
        the black-box (GameNet) vs.\ interpretable (AutoBM) contrast for 3.2.4.}
  \item \textbf{Inverting behavior generators (invert).} With $G_\phi$ fixed,
        recover $z$ from observed actions. (a) Differentiable equilibrium layers:
        Ling et al.\ embed a regularized equilibrium solver ($\mathrm{QRE}_\tau$)
        as a differentiable behavior map and back-propagate through the
        equilibrium to learn payoffs, chance, and game parameters end-to-end. (b)
        Weak-rationality private-information inference: BLUE derives feasible
        valuation intervals from ``no strictly dominated strategy,'' then learns
        history-dependent behavior with an RNN, avoiding full-equilibrium
        assumptions. (c) Social-relation / ToM inversion: composable team
        hierarchies over stochastic games recover hidden relations through
        Bayesian inverse planning~\cite{Ling2018,CuiYu2023,Shum2019}.
        \cmt[caveat]{Identification is always \emph{relative to} the assumed
        response map. Shum et al.\ use Bayesian inverse planning, not
        gradient-based neural inversion; do not imply trainable gradient
        inversion.}
  \item \textbf{Expanding the behavior-generator space (construct, II).} When the
        existing theory library is too small to contain the true generator,
        enlarge the candidate family. MARL (MAGAIL) supplies executable
        multi-agent strategic dynamics through generalized inverse RL; LLMs
        (AutoBM) generate, evaluate, and refine interpretable symbolic behavioral
        models from real human data, compiling natural-language specifications
        into PyTorch, gradient-fitting tunable parameters, and selecting by
        validation NLL through evolution plus LLM self-critique, and they
        consistently beat hand-crafted models (QRE/IA/ERC) and LLM baselines on
        ultimatum, repeated RPS, and continuous double
        auction~\cite{Song2018,AutoBM2026}.
        \cmt[caveat]{Asymmetric caveat: MARL rollouts are simulated trajectories
        that need grounding against real data; AutoBM's LLM hypotheses are not
        automatically true, but it operationalizes empirical selection (fit +
        validation NLL), so its outputs are stronger than a blanket ``synthetic
        $\neq$ evidence.'' This closes the black-box $\to$ interpretable line from
        3.2.2.}
\end{itemize}

\textbf{Story chain.} Two underidentifications (falsifiability / recovery) $\to$
neuralized priors (construct the generator's form) $\to$ generator inversion
(recover $z$) $\to$ MARL/LLM expansion of the generator space (machine
discovery, black-box $\to$ interpretable).
\end{drafting}


\subsection{Mechanism Design: Embedding Incentives into Neural Institutions}
\label{subsec:mechanism-design}

\begin{drafting}
\emph{Leading question: incentive compatibility (IC) is a counterfactual
constraint. Truthful reporting must dominate every misreport, for every type and
every opponent profile: $u_i(v_i,v_i,b_{-i}) \ge u_i(v_i,b'_i,b_{-i})\ \forall
i,v_i,b'_i,b_{-i}$. Where should IC live --- in the optimizer, the loss, the
architecture, the mechanism representation, or a post-hoc verifier?}

\textbf{Thesis.} Differentiable mechanism design advances by moving the IC
constraint step by step inside the machine: first an external constraint, then a
regret-based loss, then an architectural inductive bias, then a
truthful-by-construction representation, and finally a verifiable, interpretable,
compilable institutional object. Each step relocates IC closer to a structural
guarantee and away from an optimized penalty.

\begin{itemize}
  \item \textbf{IC as the core obstacle (AMD as pre-history).} Mechanism design
        departs from ordinary learning at one point: once published, a mechanism
        is something agents react to strategically. IC is therefore a global
        counterfactual constraint over the whole mechanism, and it cannot be
        reduced to a sample-level error. Early automated mechanism design wrote
        this as an LP / constrained optimization (variables: allocation and
        payment per type profile; constraints: IC, IR, feasibility), but
        type-space discretization explodes and injects IC
        error~\cite{CurryAMDSurvey2025}.
        \cmt[insight]{AMD's first contribution was turning mechanism design from
        theorem-proving into a computational object, before any neural network
        entered. Transition: the question becomes \emph{where} to place IC inside
        a learning system, not merely how to maximize revenue.}
  \item \textbf{IC as loss: RegretNet.} RegretNet parameterizes allocation and
        payment as a neural network that maximizes expected revenue while
        minimizing ex-post regret, trained by an augmented Lagrangian; IC is
        rewritten as ``expected ex-post regret $=0$.'' This is the first time IC
        enters a deep-learning pipeline as a differentiable
        loss~\cite{Duetting2019}.
        \cmt[tension]{Low empirical regret does not equal exact
        strategy-proofness: this remains approximate IC, useful for benchmark
        discovery and not yet sufficient for deployment. The gap drives
        everything after.}
  \item \textbf{IC as learnable violation: scaling the regret paradigm.} A wave
        of work keeps ``IC as regret penalty'' and patches four practical issues:
        better regret optimization (ALGNet casts auction learning as an
        auctioneer-vs-misreporter two-player game), better architecture
        (RegretFormer adds attention and an explicit regret budget; EquivariantNet
        bakes in bidder/item symmetry), context and variable scale (CITransNet
        handles bidder/item context with permutation-equivariant transformers),
        and objectives beyond revenue (PreferenceNet encodes fairness and
        diversity as exemplar-based
        constraints)~\cite{Rahme2021Auction,Ivanov2022,Rahme2021Equivariant,Duan2022,Peri2021}.
        \cmt[insight]{These enrich the paradigm but remain ``IC as optimized
        violation'': still approximate, not yet a structural guarantee.}
  \item \textbf{IC by construction: menus and affine maximizers.} The search
        space is restricted to natively truthful families instead of penalizing
        violations. Menu-based mechanisms (MenuNet / RochetNet) give
        truthful-by-construction single-bidder mechanisms through bidder-optimal
        menu selection; affine-maximizer auctions (AMenuNet) inherit DSIC and IR
        from AMA structure while a neural network generates AMA parameters and
        candidate allocations for scalability; GemNet pushes exact
        strategy-proofness from single- to multi-bidder by defining and enforcing
        \emph{menu compatibility} (training penalty plus post-hoc price
        transformation, discretization, and Lipschitz reasoning over the whole
        type space)~\cite{Shen2019,Duetting2019,Duan2023,Duetting2024}.
        \cmt[insight]{The defining post-RegretNet trend is the return of economic
        structure: networks search, mechanism theory bounds the search space, and
        verification/post-processing pushes approximate learning toward exact
        guarantees. \emph{Consider a small table here: IC location (loss /
        architecture bias / truthful representation / structure+verification)
        $\times$ representative line $\times$ benefit $\times$ cost.}}
  \item \textbf{IC as certificate: toward verified institutional generation.}
        Once IC is structurally represented, the question turns institutional:
        can the mechanism be verified, interpreted, and deployed as a rule?
        BundleFlow pushes DSIC menu learning into combinatorial auctions through
        ODE/flow-generated bundle distributions, scaling to hundreds of items;
        the discretization-free truthful-learning frontier learns truthful
        mechanisms over continuous outcome spaces via affine-maximizer structure;
        and LLM-AMD reformulates mechanism design as code generation plus a
        problem-specific fixing process for feasibility and strategy-proofness,
        rediscovering Myerson and reserve-price/redistribution mechanisms and
        explaining neural solutions by
        programming-by-example~\cite{WangBundleFlow2025,TEDI2025,LiuGuoConitzer2025}.
        \cmt[insight]{The endpoint is a search + structure + verification +
        explanation pipeline, not a black-box neural auction. Position LLMs
        precisely: they act as an institutional compiler, not as another
        RegretNet.}
  \item \cmt[todo]{Add the ``Where does IC live?'' narrative figure (external
        constraint $\to$ differentiable loss $\to$ better regret learning $\to$
        truthful representation $\to$ structure + verification $\to$
        code/certificate) and write the opening and closing paragraphs.}
\end{itemize}
\end{drafting}