% conclusion.tex
% Section 5: Synthesis and Outlook
% Revised against review_r1.md (Section 5 row: R2' core, R4', R3', R1').
% Same house style as the other sections: the body is wrapped in a `drafting`
% block (see review.tex) that renders in blue under \reviewtrue and is hidden
% under \reviewfalse. Writing plan / guardrails / positioning are marked \cmt.
%
% Requires the main preamble to contain \input{review}.

\section{Synthesis and Outlook}
\label{sec:synthesis}

\begin{drafting}
\textbf{Section framing.} The conclusion does the one thing the body could not:
read the six domains as a single movement and say what that movement implies
going forward. It closes the loop opened in Section~\ref{sec:intro} (the
universally quantified certificate confronting procedures with only finite
access) and Section~\ref{sec:recap} (game theory ending on computation, deep
learning ending on construction at scale).
\cmt[skeleton]{Revised skeleton per the updated review mapping (R1$'$, R2$'$,
R3$'$, R4$'$; R2$'$ is the main landing of the whole paper). Seven movements:
(1) synthesis under the revised refrain --- restate the shared problem as
``universal certificate vs.\ finite access'' and read
Section~\ref{sec:ai4gt} as the query regime, Section~\ref{sec:gt4ai} as the
observation regime, with 4.1/4.2 marked as the intermediate/verification cases;
(2) the traveling object restated per R1$'$: what crosses the boundary is the
universally quantified certificate, and the two directions of travel are the
two access regimes; (3) R2$'$ MAIN LANDING: name and define the checkable
strategic certificate (construction--verification cost decoupling), anchor it
in 2.1's PPAD prototype, give the four-class taxonomy
(Table~\ref{tab:certificates}), and close the three-node circuit specification
$\to$ construction $\to$ verification; (4) co-evolution --- tightened
descendants of the two intro paragraphs, now run through that circuit; (5)
R3$'$: the extensional capability table (Table~\ref{tab:capabilities}),
supplied by 2.2's five movements, plus the row-mode LLM observation in
artifact-agnostic wording; (6) research directions, former items 1 and 6
merged under R2$'$; (7) close on the fourth turn (R4$'$) as a corollary of the
certificate definition, explicitly distinguished from the third turn. This is
the third and final occurrence of the merged R2$'$+R4$'$ claim; the Abstract
and Section~\ref{sec:intro} carry the first two and must be revised in step.
Keep the section shorter than the body sections.}

\cmt[step]{(1) Synthesize. Restate the shared problem in the revised refrain
(R1$'$), then assign each direction its access regime. Reuse Part~I's
recurring-move sentence and Part~II's scalar$\to$object spine in substance; the
regime language wraps them, it does not replace them. End on the verification
obstruction so movement (3) is earned.}
The six domains answer one question from two sides: how can a strategic
property that quantifies universally over counterfactual interaction, over
deviations not taken, types not reported, and opponents not encountered, be
certified by a procedure that has only finite access to the game? The two parts
of the survey differ in the form that finite access takes.

Section~\ref{sec:ai4gt} operates in the \emph{query} regime. A counterfactual
can be computed on demand: a deviation can be simulated, a misreport can be
priced, a best response can be trained. Each query costs computation, and the
quantified space outgrows any budget, so the obstruction is computational.
Within this regime the recurring move relocates the game-theoretic constraint
into a computational substrate (a differentiable residual, a no-regret
trajectory, an architectural bias, or a checkable certificate), while the
guarantee re-enters through a classical layer such as linear programming,
no-regret theory, safe resolving, or symbolic proof.

Section~\ref{sec:gt4ai} operates in the \emph{observation} regime. A
counterfactual can no longer be queried: a deployed opponent, a human rater, or
a rival's pricing algorithm reveals only the play it happens to produce, so the
obstruction is epistemic. Two of its settings sit between the poles: strategic
evaluation holds query access to the agent under test while a finite population
stands in for the true opponents
(Section~\ref{subsec:strategic-evaluation}), and debate-style training must
certify a stronger model from finite interaction with a weaker judge
(Section~\ref{subsec:training-alignment}). Within this regime the recurring
move replaces a scalar with a game-theoretic object: a benchmark score with
exploitability, a fixed reward with the equilibrium of a game, source-code
inspection with an equilibrium-level audit. In both regimes the residual demand
is the same, and it is the verification obstruction of
Section~\ref{sec:intro}: a universal claim must be certified from finite
access.

\cmt[step]{(2) The common movement, restated per R1$'$. The traveling object is
the universally quantified certificate; the two directions of travel are the
two access regimes just distinguished. The observation is the synthesis, not a
restatement of (1).}
A single object makes both journeys: the universally quantified certificate.
Section~\ref{sec:ai4gt} carries it from specification toward construction,
building or approximating under query access the certificate that classical
theory could only define. Section~\ref{sec:gt4ai} carries it the opposite way,
attaching a counterfactual specification to systems built before anyone asked
what they should satisfy, and testing that specification against the behavior
the systems emit. The survey addresses the specification--construction gap named
in Section~\ref{sec:intro} from both directions, using procedures whose finite
access takes the two forms above.

\cmt[step]{(3) R2$'$ main landing. Order is load-bearing: definition
(construction--verification decoupling) $\to$ PPAD prototype from 2.1 $\to$
taxonomy table $\to$ three-node circuit. Each later element presupposes the
earlier one. Research-direction items 1 and 6 merge under this concept in
movement (6).}
Both journeys end at the same requirement, and meeting it is what answers the
verification obstruction: the certificate must be \emph{checkable}, meaning its
verification cost is decoupled from its construction cost. Finding an object
that satisfies a universally quantified strategic property may be expensive;
checking a proposed object against a stated trust model must be cheap.
Section~\ref{subsec:recap-gt} closes on the prototype of this asymmetry: a Nash
equilibrium is guaranteed to exist and appears intractable to find, yet a
candidate profile is verified in polynomial
time~\cite{Papadimitriou1994,DaskalakisGoldbergPapadimitriou2009}. The six
domains widen that asymmetry from a complexity class into an engineering
discipline, and their material sorts into four classes of checkable strategic
certificate, recorded with their quantified objects, access requirements, and
trust models in Table~\ref{tab:certificates}: \emph{symbolic proof}, in which a
machine-checkable derivation accompanies the
solution~\cite{PrimeNash2025,LegoNE2025}; \emph{structural guarantee}
(\emph{construction-is-truth}), in which membership in a representation family
makes the property hold by construction~\cite{Shen2019,Duan2023,Duetting2024};
\emph{protocol guarantee},
in which the rules of an interaction guarantee the honest party a winning
strategy~\cite{Irving2018,BrownCohen2023}; and \emph{behavioral statistical
test}, in which finite observed play bounds the counterfactual property under
the weakest trust
assumptions~\cite{HartlineLongZhang2024,HartlineWangZhang2025}.

\begin{table}[t]
\centering
\footnotesize
\caption{Four classes of checkable strategic certificate. Each class decouples
verification cost from construction cost; they differ in what the certificate
quantifies over, in the access the verifier needs to the certified object, and
in what the verifier must trust.}
\label{tab:certificates}
\begin{tabular}{p{0.13\textwidth}p{0.21\textwidth}p{0.16\textwidth}p{0.21\textwidth}p{0.15\textwidth}}
\hline
Class & Quantifies over & Access to object & Verifier trusts & Evidence \\
\hline
Symbolic proof & all deviations in an explicit game; all inputs of a
synthesized algorithm & white-box (game or algorithm description) & a proof
checker & \S\ref{subsec:constructive-equilibrium}:
\cite{PrimeNash2025,LegoNE2025} \\
Structural guarantee (construction-is-truth) & all types and all misreports & white-box (mechanism
representation) & the representation family, plus post-hoc verification & \S\ref{subsec:mechanism-design}:
\cite{Shen2019,Duan2023,Duetting2024} \\
Protocol guarantee & all strategies of a dishonest prover & black-box model
inside a white-box protocol & the protocol, its complexity assumptions, and the
judge & \S\ref{subsec:training-alignment}:
\cite{Irving2018,BrownCohen2023} \\
Behavioral statistical test & all deviations available to the audited algorithm
& black-box (observed play only) & a concentration bound on finite data
(weakest) & \S\ref{subsec:deployment-governance}:
\cite{HartlineLongZhang2024,HartlineWangZhang2025} \\
\hline
\end{tabular}
\end{table}
\cmt[todo]{Table format: plain \texttt{hline} rules because the current
preamble loads no \texttt{booktabs}; switch to
\texttt{toprule/midrule/bottomrule} if the package is added. Confirm column
wording of ``trust model'' with co-authors; the row order follows the order in
which the survey introduces each class.}

The four classes close the two directions into a circuit with three nodes. A
specification is made constructible (Section~\ref{sec:ai4gt}); the construction
is checked against its specification through one of the four certificate
classes; and what the check reveals (an exploitability gap, an incoherent
preference, a failed regret test) becomes the revised specification that the
next construction must meet (Section~\ref{sec:gt4ai}). Verification is the node
at which the other two trust each other: it keeps construction accountable to
specification, and it keeps specification answerable to what can actually be
built.

\cmt[step]{(4) Co-evolution. Tightened descendants of the two paragraphs cut
from the intro; the intro's \cmt[todo] points here. The circuit of movement (3)
now carries the loop, and the closing sentence develops the intro's single
forward-pointing sentence with the certificate as the enabling condition.}
Read through this circuit, the domains describe a co-evolution. On the
specification-to-construction arc, strategic specifications become executable:
a Nash equilibrium, an incentive-compatible mechanism, or a
correlated-equilibrium bound stops being only an analytic object and becomes
something a learning system can search for, approximate, and certify. The
game-theoretic constraint acquires a constructive arm it lacked once the
relevant games grew large.

On the construction-to-specification arc, executions generate their own demand
for specification. Once a trained system negotiates, prices, or persuades, its
competence becomes opponent-relative and its alignment becomes a relation among
policies, so the system itself calls for a strategic property to evaluate and
govern. Construction acquires a specification arm it lacked when behavior was
scored against a fixed test set. The loop announced in
Section~\ref{sec:intro} therefore runs through the verification node: more
capable strategic AI demands richer specifications to stay legible and
governable, and richer specifications, once their certificates are checkable,
unlock the construction of still more capable systems.

\cmt[step]{(5) R3$'$. Where learning enters the circuit: the extensional
capability table, whose rows are supplied by
Section~\ref{subsec:recap-dl}'s arc, then the row-mode observation about
symbolic generation. Guardrail (per review): do NOT claim LLMs are the first
systems to be both constructor and subject --- self-play RL already was
(AlphaStar is evaluated by the Nash of its own league). The claim is
specification-side participation; its reason is the medium; the wording stays
artifact-agnostic (``learners that work in the written medium of
specifications''), with LLMs named as the current instance.}
Learning enters this circuit through four capabilities, each traceable to the
arc of Section~\ref{subsec:recap-dl} and each load-bearing in specific cells of
the survey (Table~\ref{tab:capabilities}): \emph{gradient residualization},
which rewrites a fixed-point condition as a loss; \emph{no-regret dynamics},
which relax a one-shot universal test into a property of a learning trajectory;
\emph{population search}, which approximates an intractable strategy space by a
growing empirical one; and \emph{symbolic generation}, which emits programs,
models, and proofs in the notation the theory itself uses. The first three
capabilities construct or approximate, and in every cell they occupy the
certificate is restored by a classical layer: linear programming and no-regret
theory in Section~\ref{subsec:constructive-equilibrium}, mechanism-theoretic
structure in Section~\ref{subsec:mechanism-design}, protocol design and
statistical testing in Sections~\ref{subsec:training-alignment}
and~\ref{subsec:deployment-governance}.

\begin{table}[t]
\centering
\footnotesize
\caption{Where the four learning capabilities bear load across the six domains.
Rows originate in the developmental arc of Section~\ref{subsec:recap-dl};
cells name the instantiation that carries the argument in each part.}
\label{tab:capabilities}
\begin{tabular}{p{0.14\textwidth}p{0.20\textwidth}p{0.27\textwidth}p{0.25\textwidth}}
\hline
Capability & Origin (\S\ref{subsec:recap-dl}) & Specification $\to$
construction (\S\ref{sec:ai4gt}) & Construction $\to$ specification
(\S\ref{sec:gt4ai}) \\
\hline
Gradient residualization & the network as a differentiable parameterized map &
regularized best-response residuals
(\S\ref{subsec:constructive-equilibrium})~\cite{NashPG2025,LiuFarinaOzdaglar2025};
differentiable equilibrium layers
(\S\ref{subsec:strategic-inference})~\cite{Ling2018}; regret-as-loss
(\S\ref{subsec:mechanism-design})~\cite{Duetting2019} & saddle-point
adversarial training
(\S\ref{subsec:training-alignment})~\cite{Goodfellow2014,Madry2018} \\
No-regret dynamics & training as an iterative update trajectory & CFR and Deep
CFR
(\S\ref{subsec:constructive-equilibrium})~\cite{Zinkevich2007,BrownDeepCFR2019}
& preference-equilibrium alignment
(\S\ref{subsec:training-alignment})~\cite{Munos2024,Zhang2025INPO}; the audited
rationality property
(\S\ref{subsec:deployment-governance})~\cite{Hartline2026} \\
Population search & scale and self-play & EGTA and PSRO
(\S\ref{subsec:constructive-equilibrium})~\cite{WellmanTuylsGreenwald2025,McAleerPipelinePSRO2020}
& Nash averaging, $\alpha$-Rank, league evaluation
(\S\ref{subsec:strategic-evaluation})~\cite{Balduzzi2018,Omidshafiei2019,Vinyals2019} \\
Symbolic generation & code and structured text generated at scale & equilibrium
proofs and algorithm synthesis
(\S\ref{subsec:constructive-equilibrium})~\cite{PrimeNash2025,LegoNE2025};
interpretable behavior models
(\S\ref{subsec:strategic-inference})~\cite{AutoBM2026}; mechanism-as-code
(\S\ref{subsec:mechanism-design})~\cite{LiuGuoConitzer2025} & the subject under
evaluation, alignment, and governance
(\S\ref{subsec:strategic-evaluation}--\ref{subsec:deployment-governance})~\cite{GTBench2024,Akata2025} \\
\hline
\end{tabular}
\end{table}

The fourth row behaves differently, and the difference shows when the row is
read as a whole. Its Section~\ref{sec:ai4gt} cells return specification-side
objects: a machine-checkable equilibrium proof~\cite{PrimeNash2025}, a
synthesized algorithm with a provable approximation bound~\cite{LegoNE2025}, an
interpretable symbolic model of human play~\cite{AutoBM2026}, a mechanism
delivered as code with its incentive properties repaired and
explained~\cite{LiuGuoConitzer2025}. These are the cells in which the
certificate re-enters whole; every other row leaves its restoration to a
classical layer. The same systems appear throughout Section~\ref{sec:gt4ai}
purely as subjects, evaluated, aligned, and
governed~\cite{GTBench2024,Akata2025}. Self-play systems already occupied both
of those roles: AlphaStar constructs a league and is evaluated by the Nash of
that league~\cite{Vinyals2019}, so the pairing of constructor and subject
predates this row. What the row adds is operation on a third side, the
specification itself: these learners produce and consume the proofs,
mechanisms, and audit rules, because they work in the written medium in which
specifications are expressed. The observation attaches to that capability,
whatever artifact currently hosts it.

\cmt[step]{(6) Research directions. Former items 1 and 6 merge under R2$'$
(both were ``learned construction + checkable certificate''); the remaining
items keep their anchors in the body sections and are recast in the
certificate/regime vocabulary where it sharpens them. All flagged tentative.}
Several open problems follow from the synthesis; the list is a first pass, and
each item is flagged tentative pending the authors' final scoping.
\begin{itemize}
  \item \textbf{Learned construction with machine-checkable certificates beyond
        two-player zero-sum.} The complexity partition that organizes
        Section~\ref{subsec:constructive-equilibrium} (linear-programming
        tractability in zero-sum, PPAD-hardness in general-sum) still bounds
        what learned solvers can guarantee, and the methods that scale retreat
        to (C)CE and lose the exact certificate. Symbolic generation restores
        the full certificate, so far on instances and algorithm families of
        modest size~\cite{PrimeNash2025,LegoNE2025,LiuGuoConitzer2025}. Whether
        a learned constructor can emit general-sum solutions whose certificates
        are machine-checkable at scale, and under what verification the emitted
        certificates deserve trust, is a single question that joins the
        survey's construction frontier to its certification frontier.
  \item \textbf{An identifiability theory for neural behavior generators.}
        Section~\ref{subsec:strategic-inference} is the observation regime at
        its purest: fit is cheap and identification hard, with recovery
        guaranteed only under auxiliary assumptions such as entropy
        regularization~\cite{FreihautRamponi2025}. When the latent strategic
        state is recoverable from observed behavior, and relative to which
        assumed response map, remains open.
  \item \textbf{Exact incentive compatibility at deployment scale.}
        Section~\ref{subsec:mechanism-design} reached the structural-guarantee
        class, but exact strategy-proofness over multi-bidder, combinatorial,
        or continuous type spaces is enforced only
        partially~\cite{Duetting2024,WangBundleFlow2025,TEDI2025}. Mechanisms
        that are truthful by construction and verified end-to-end at scale are
        still to be built.
  \item \textbf{Preference coherence as a precondition for alignment.}
        Section~\ref{subsec:strategic-evaluation} noted that exploitability and
        rankings presuppose a coherent utility whose existence is itself
        uncertain for current models~\cite{RossKimLo2024,Mazeika2025}. Auditing
        whether the utility exists and selecting the equilibrium it should play
        are treated separately by current pipelines.
  \item \textbf{Equilibrium-level governance standards.}
        Section~\ref{subsec:deployment-governance} presented calibrated-regret
        auditing as a recent program in the behavioral-test
        class~\cite{HartlineWangZhang2025,Hartline2026}. Turning it into a
        deployable standard, with an agreed equilibrium-level property, test,
        and sample-complexity budget, is a direction for both theory and
        policy.
\end{itemize}
\cmt[tentative]{All five are first-pass candidates --- confirm scope and
pruning with co-authors. Former items 1 and 6 are merged above per the review.
The previously contemplated cross-cutting item (a common formal language for ``a
counterfactual strategic property represented inside a computational system'')
is no longer listed: the checkable-certificate definition and
Table~\ref{tab:certificates} now supply exactly that language, so the item
would duplicate movement (3). Decide whether the list stays prose-itemized or
moves to a short table.}

\cmt[step]{(7) Close on the future of game theory itself, matching the intro's
promise. R4$'$: the fourth turn appears as a corollary of the certificate
definition (its third occurrence after the Abstract and
Section~\ref{sec:intro}), and the distinction from the third turn is written
out so the turn does not collapse back into AGT for a TCS reader. The
distinguishing data are already on the page: the access and trust columns of
Table~\ref{tab:certificates}.}
Game theory has turned its binding question three times: from whether an
equilibrium exists, to how an institution should be designed, to what an
equilibrium costs to compute (Section~\ref{sec:recap}). The certificate
definition makes a fourth turn a corollary. Once strategic guarantees are
required to be checkable, and once the systems that must carry them are
learning systems with finite access, the binding question becomes whether a
strategic guarantee can be constructed and verified inside a deployed learning
system. The fourth turn differs from the third in its certified object and in
its starting data. Algorithmic game theory certifies properties of an
explicitly given game: the description is the input, full query access comes
with it, and the open question is the cost of computation. The fourth turn
certifies deployed learning systems and the empirical games they induce: the
description is unavailable, and the test must start from structure (an
architecture, a protocol) or from behavior (observed play), as the access and
trust columns of Table~\ref{tab:certificates} record. Game theory thereby gains
a constructive and a governance face alongside its analytic one, and the
systems it governs gain a vocabulary precise enough to stay accountable as they
grow more capable.
\end{drafting}
