% gt4ai.tex
% Section 4: Game Theory for Strategic AI
% Preliminary English draft outline transcribed from the section-4 planning note.
% Each subsection body is wrapped in a `drafting` block (see review.tex): it
% renders in blue under \reviewtrue and is hidden under \reviewfalse. Self-notes
% / guardrails / positioning are marked with \cmt.
%
% Requires the main preamble to contain \input{review}.

\section{Game Theory for Strategic AI}
\label{sec:gt4ai}

\begin{drafting}
\textbf{Part II framing.} Static ML asks whether a model scores well; strategic
AI asks whether a policy stays robust when other agents adapt. \emph{When AI
systems act strategically alongside other adaptive agents, what should be
evaluated, optimized, and governed?} The structure follows evaluation $\to$
equilibrium-seeking training $\to$ deployment governance
(\ref{subsec:strategic-evaluation}, \ref{subsec:training-alignment},
\ref{subsec:deployment-governance}).
\cmt[spine]{The spine across all three subsections: each replaces a \emph{scalar}
with a \emph{game-theoretic object}. Benchmark score $\to$
exploitability/equilibrium; fixed scalar reward $\to$ equilibrium selection;
source-code inspection $\to$ equilibrium-level audit.}
\end{drafting}


\subsection{Strategic Evaluation: Competence Is Opponent-Relative}
\label{subsec:strategic-evaluation}

\begin{drafting}
\emph{Leading question: if strategic competence depends on opponents, protocols,
and strategy spaces, can it still be evaluated as a scalar model score?}

\textbf{Thesis.} The first role of game theory is diagnostic. It asks what it
means to evaluate competence when performance is opponent-relative, and, prior to
that, whether the agent even holds a coherent utility for any such evaluation to
be well-posed. The unit of evaluation shifts from an isolated model score to a
policy under interaction. In the article's refrain, strategic evaluation is an
intermediate access regime: the evaluator can query the policy under test by
running new interactions, but deployment opponents, protocols, and populations
are represented only by a finite surrogate population. Exploitability is the
cleanest example. It compresses a universally quantified certificate, the maximum
gain available to any best-responding deviation, into a scalar gap that can only
be approximated from finitely many queried opponents.

\begin{itemize}
  \item \textbf{Step 1 --- static benchmarks fail under adaptive opponents.}
        Accuracy and win-rate stop being meaningful once opponents adapt. The
        canonical RL-side diagnosis is that proliferating benchmark suites and
        baselines dilute evaluation and invite cherry-picking, which motivates a
        population-level unit of evaluation~\cite{Balduzzi2018}.
  \item \textbf{Step 2 --- opponent-relative competence: the RL/EGTA vocabulary.}
        This line \emph{predates and grounds} today's LLM-evaluation vocabulary.
        In conceptual order: EGTA names population-level evaluation by estimating
        an empirical game from simulation or data and inducing a verdict through a
        solution concept~\cite{Wellman2006}; PSRO with joint-policy correlation
        formalizes overfitting-to-training-opponents, i.e.\ the profile
        dependence of competence~\cite{Lanctot2017}; Nash averaging reweights a
        meta-game by its max-entropy Nash to discount redundant tasks and weak
        baselines~\cite{Balduzzi2018}; $\alpha$-Rank tracks non-transitivity and
        sinks in a population ranking~\cite{Omidshafiei2019}; open-ended learning
        separates transitive from non-transitive structure (in a cycle one can
        ``get stronger but against whom is unclear''
        )~\cite{Balduzzi2019Open}; and Spinning Tops raises non-transitivity from
        a toy example to the generic geometry of real-world
        games~\cite{Czarnecki2020}. The common move is to replace a fixed test
        set by a finite empirical population that stands in for the unbounded
        opponent space against which the universal no-profitable-deviation claim
        is meant to range.
        \cmt[caveat]{Flag $\alpha$-Rank's scalability limit (Yang et al.\ 2020,
        \cite{Yang2020}) at the point of mention.}
        \cmt[evidence]{Engineering evidence that opponent-relative robustness is
        already the deployed norm: AlphaStar outputs a Nash distribution over
        least-exploitable agents~\cite{Vinyals2019}; DeepStack / Libratus /
        Pluribus report exploitability (LBR) as the core metric instead of
        win-rate~\cite{Moravcik2017,BrownSandholm2018,Brown2019Pluribus}.}
  \item \textbf{Step 3 --- the same vocabulary, latest chapter on LLMs.}
        Migrated to LLM strategic behavior, the vocabulary exposes profile
        dependence: GTBench shows strategic reasoning varies with game
        structure~\cite{GTBench2024}; repeated-games studies show that
        cooperation and coordination are modulated by prompt, opponent, and
        interaction (GPT-4 is unforgiving in the Prisoner's Dilemma yet cannot
        coordinate with an alternating opponent in Battle of the
        Sexes)~\cite{Akata2025}. These results are subject-side evidence: the
        language model is the policy being evaluated as a strategic actor, and
        its apparent competence changes with the game, opponent, and interaction
        protocol. LLM benchmarks are a downstream application of the EGTA /
        $\alpha$-Rank tradition, not an island.
  \item \textbf{Step 4 --- precondition: preference coherence is the ground
        floor.} Steps 1--3 assume the agent holds a well-defined utility to play
        from; equilibrium, best response, and payoff all presuppose it. When
        preferences are incoherent or drift with framing, exploitability and
        rankings rest on sand. Utility-theory work audits this precondition.
        ``LLM economicus?'' tests inequity/risk-loss aversion and time
        discounting and finds current LLMs neither fully human-like nor fully
        economicus, struggling to stay consistent across
        settings~\cite{RossKimLo2024}; Utility Engineering fits Thurstonian
        random-utility models and finds preference coherence
        (transitivity/completeness) \emph{increasing with model
        scale}~\cite{Mazeika2025}.
        \cmt[footnote]{Footnote the behavioral-economics Turing-test line:
        \cite{Mei2024,Horton2023,Aher2023}.}
        \cmt[framing]{Ross--Kim--Lo include a competence test and analyze only
        models that pass it, itself an admission of the split between prior
        preference coherence and posterior strategic competence. Steps 2--3 audit
        play given a utility; Step 4 audits whether the utility exists.}
\end{itemize}
\end{drafting}


\subsection{Training and Tuning: Alignment as Equilibrium Selection}
\label{subsec:training-alignment}

\begin{drafting}
\emph{Leading question: when the alignment signal cannot be reduced to a fixed
scalar objective, because it is defined only against an adversary, only as a
non-transitive preference, or only as what a bounded verifier can be brought to
accept, what should training optimize instead?}

\textbf{Thesis.} The access regime is verification-dominated: the trainer can
stage interactions with adversaries, preference models, or judges, but the
property to be certified ranges over responses and arguments that finite
training cannot enumerate. The unified answer replaces ``maximize a fixed scalar
objective'' with ``select the equilibrium of a game.'' Two preliminaries set up
the strands. First, self-play is the shared computational mechanism of all three
strands, not a classification axis; its ``self-play $\to$ Nash'' foundation is
fictitious play and its deep
variants~\cite{Brown1951,Robinson1951,Heinrich2015,HeinrichSilver2016}. Second,
the three strands are distinguished by \emph{what the equilibrium certifies}:
robustness (Strand 2), preference (Strand 1), truth (Strand 3). We present them
in the order 2 $\to$ 1 $\to$ 3, because GAN is the historical template for
``training as equilibrium-seeking'' that the other two inherit. For LLM
alignment, the learned model is the strategic subject being aligned: training
builds games around its adversarial behavior, preference responses, or
argumentative conduct.

\textbf{Strand 2 --- Adversarial games (template + dynamics diagnosis).}
\begin{itemize}
  \item Template: the GAN as zero-sum minimax, with Nash where the discriminator
        cannot distinguish real from generated data~\cite{Goodfellow2014}.
  \item Alignment instances: robust training as a saddle point~\cite{Madry2018};
        red-teaming as an attacker--defender game~\cite{Perez2022}.
  \item Difficulty (empirical and conceptual): the fence drawn in
        Section~\ref{subsec:recap-gt} matters here. Reciprocal consistency, even
        in the local Nash sense, is a static certificate; dynamical stability and
        stable fixed points are properties of a chosen update rule. First-order
        methods can rotate around a local Nash in the last iterate, fail to
        converge, or face GAN games with no local Nash
        equilibrium~\cite{FarniaOzdaglar2020}.
  \item \cmt[box]{Boxed supplement --- differential game dynamics: decompose the
        game Jacobian into symmetric (potential) and antisymmetric (Hamiltonian)
        parts and use it to design SGA toward stable fixed
        points~\cite{BalduzziSGA2018,Letcher2019}. Position it NOT as a Nash
        solver or core anchor but as the dynamical lens for \emph{why}
        equilibrium-seeking training rotates / fails to converge; it
        distinguishes reciprocal consistency / local Nash from dynamical
        stability / stable fixed point, and explains last-iterate cycling.
        Adjacent: \cite{Mescheder2017, Daskalakis2017}.}
\end{itemize}

\textbf{Strand 1 --- Preference games (template for ``preference is not scalar'').}
\begin{itemize}
  \item Insufficiency: Bradley--Terry forces transitivity, while real and
        aggregated preferences can be intransitive and
        cyclic~\cite{BradleyTerry1952,May1954,Tversky1969,Dudik2015}.
  \item Conceptual anchor (NLHF): treat the pairwise preference model as a
        symmetric two-player game, solve for a regularized Nash equilibrium, and
        converge through last-iterate Nash-MD, so that RLHF becomes equilibrium
        selection~\cite{Munos2024}.
  \item Scalable implementations (lead with INPO): INPO approximates Nash by
        no-regret self-play with a single loss~\cite{Zhang2025INPO}; DNO is an
        iterative-DPO variant~\cite{Rosset2024DNO}; SPPO uses a constant-sum
        Hedge update~\cite{Wu2024SPPO}; SPO/Minimaximalist is the social-choice
        Minimax-Winner template~\cite{Swamy2024SPO}.
  \item Back to the box (convergence dynamics): mirror descent gives only
        average-iterate convergence while the last iterate orbits
        Nash~\cite{Mertikopoulos2018}; last-iterate fixes use extragradient-style
        updates.
  \item PSRO bridge (one line): a robust policy is a robustness search over a
        strategy population~\cite{Lanctot2017}.
        \cmt[footnote]{Distribution-matching self-play: SPIN, \cite{Chen2024SPIN}.}
\end{itemize}

\textbf{Strand 3 --- Interactive-proof games (template for ``the judge is weaker
than the model'').}
\begin{itemize}
  \item Complexity foundation: debate turns the verification obstruction into a
        training game. The weak judge sees only a finite interaction, but the
        guarantee quantifies over the full space of arguments a dishonest prover
        might construct. Self-play on a zero-sum debate game with a
        polynomial-time judge can reach optimal play in PSPACE, isomorphic to
        IP$=$PSPACE~\cite{Irving2018}.
  \item Two canonical protocols: Debate (symmetric: two provers + judge); and
        Prover--Verifier Games (asymmetric, fitting weak-to-strong, where only
        some variants provably have ideal equilibria and sequential forms can
        fail)~\cite{Anil2021}, scaled to legibility
        training~\cite{Kirchner2024}.
  \item Obstruction and refinement: obfuscated arguments let a dishonest prover
        construct claims that take exponential time to
        refute~\cite{BarnesChristiano2020}; doubly-efficient debate is the
        checkable-certificate taxonomy's protocol-guarantee case, because the
        protocol itself guarantees the honest prover a polynomial-time winning
        strategy while keeping verification bounded~\cite{BrownCohen2023}.
  \item Empirical landing: with stronger expert debaters and weaker non-expert
        judges, debate consistently raises accuracy (76\%/88\% vs.\ 48\%/60\%
        baselines), and adversarial protocols beat non-adversarial
        ones~\cite{Khan2024}.
  \item \cmt[positioning]{Closing position: when the supervision signal is no
        longer reliable, an interactive-proof game designs the equilibrium to
        \emph{certify truth}; the LLM remains the strong subject being aligned,
        not a generic meta layer.}
\end{itemize}
\end{drafting}


\subsection{Deployment Governance: Local Alignment Is Not System Alignment}
\label{subsec:deployment-governance}

\begin{drafting}
\emph{Leading question: when individually aligned AI agents interact through
markets, institutions, and delegation chains, what system-level equilibria do
they induce?}

\textbf{Thesis.} The third role of game theory is institutional. It asks what
system-level equilibria emerge when individually aligned agents interact, and how
those equilibria can be audited, reshaped, or governed. The question shifts from
whether a single agent is aligned to what locally aligned agents induce in
interaction. A single game-theoretic line makes the cut: the economics of
no-regret learning plus calibrated-regret auditing. This is observational access
in its pure form. The auditor has behavioral data from deployed systems, not
query access to every counterfactual deviation and not trusted source-code
inspection, yet must certify an equilibrium-level universal property from finite
observations. That is the governance face of the verification obstruction.

\begin{itemize}
  \item \textbf{Step 1 --- delegation games: a conceptual decomposition for
        ``local $\neq$ system''.} Multi-principal multi-agent settings are
        formalized as delegation games, with failure modes decomposed
        orthogonally into (control $\times$ cooperation) $\times$ (alignment
        $\times$ capabilities); no single measure suffices, all four together do,
        and a bound on principal welfare loss follows~\cite{Sourbut2024}.
        \cmt[positioning]{Position as conceptual framework / decomposition (why a
        per-agent property is not composable), not the main theoretical support
        of this subsection.}
  \item \textbf{Step 2 --- algorithmic pricing as adaptive market failure
        (phenomenon + key qualifications).} Q-learning agents in Bertrand
        oligopoly learn supra-competitive prices without communication, supported
        by punishment-then-return strategies; this is a \emph{simulation} result
        with no general convergence guarantee~\cite{Calvano2020}. Three
        qualifications follow. ``Collusion'' is largely an artifact of the
        \emph{asynchronous} learning protocol and mostly vanishes under
        synchronous learning or minimal economic structure~\cite{Asker2022}; the
        outcomes are fragile and exploitable by an optimizing
        opponent~\cite{denBoer2022}; and a Folk-Theorem-style result makes
        collusion only one of many reachable payoffs, not a
        necessity~\cite{Cartea2025}. The governance target is
        \emph{whether observed strategic behavior satisfies an equilibrium-level
        property}; source-code labels such as ``Q-learning'' leave that target
        unresolved. This shift pushes governance from source-code inspection to
        behavioral audit. These
        pricing and collusion studies are subject-side evidence for strategic
        AI: the learning pricing policy is itself the strategic market actor
        whose adaptive behavior changes the institutional outcome.
  \item \textbf{Step 3 --- calibrated-regret audit as the governance response
        (theoretical climax).} The ``for AI'' bridge is direct: classical
        algorithmic game theory predates AI agents, yet these AI learning agents
        \emph{are} precisely the no-regret learners those theorems quantify over.
        \begin{itemize}
          \item Theoretical basis (making ``observed behavior'' an auditable
                object): no-internal-regret learning converges to correlated
                equilibrium~\cite{FosterVohra1997,HartMasColell2000,BlumMansour2007};
                in smooth games the social welfare of no-regret learners is bounded
                by the price of anarchy~\cite{Roughgarden2015,SyrgkanisTardos2013};
                modern algorithms converge at near-optimal
                rates~\cite{Syrgkanis2015,Anagnostides2022,Dagan2024}.
          \item Audit mechanism: instead of inspecting source code, test whether
                a seller has vanishing (pessimistic) calibrated regret. The test
                compares realized revenue with the best calibrated deviation
                available after each observed history; small calibrated regret
                means no such history-contingent deviation would have improved
                the seller's payoff by much. This is the fourth class of
                checkable strategic certificate: a behavioral statistical test,
                black-box in its access model and weak in its trust assumptions,
                whose target is still a universal no-profitable-deviation
                property. Passing the test proves that
                (i) any no-regret algorithm can be retrofit to pass, (ii) passers
                cannot be manipulated into supra-competitive prices, and (iii) a
                sample-complexity bound
                holds~\cite{HartlineLongZhang2024,HartlineWangZhang2025}; a recent
                survey proposes no-swap-regret as a rationality axiom for
                algorithmic agents, bridging classical AGT to AI
                agents~\cite{Hartline2026}.
                \cmt[wording]{Frame this as a \emph{recent program} building on
                classical no-regret theory, not an established field. Cite the two
                CSLAW papers as primary; the 2026 survey as a programmatic
                reference.}
          \item Why calibrated regret is the right hinge: related mechanism
                results show that the same problem appears whenever rules face
                adaptive learners. A seller can exploit a no-regret buyer across
                time~\cite{Braverman2018}, and first-price bid suppression shows
                how learning dynamics can reshape bidding incentives
                themselves~\cite{BanchioSkrzypacz2022}. The audit asks whether
                such self-interested adaptation remains inside an acceptable
                equilibrium-level envelope.
        \end{itemize}
        This closes the line opened in Section~\ref{subsec:mechanism-design}.
        There the problem is to design rules so self-interested AI behavior is
        good by construction; here the problem is to audit deployed rules and
        learners from behavior alone, under the weakest access regime.
\end{itemize}
\end{drafting}
