% introduction.tex
% Section 1: Introduction
% Same house style as ai4gt.tex / gt4ai.tex: the section body is wrapped in a
% `drafting` block (see review.tex) that renders in blue under \reviewtrue and
% is hidden under \reviewfalse. Writing plan / guardrails / positioning are
% marked with \cmt.
%
% Requires the main preamble to contain \input{review}.
%
% REVISION (this pass) --- implements the Section-1 row of the corrected
% review mapping:
%   R1'      the three-obstruction taxonomy of the previous draft is promoted
%            to the article-wide refrain and re-derived as "universally
%            quantified certificate vs. finite access", with two access
%            regimes (query / observational) plus the cross-cutting
%            verification question; on/off-path language survives only as the
%            special case of the observational regime; every Part I/II bullet
%            now carries its regime tag, redeeming the refrain's promise.
%   R2'+R4'  step (5)'s forward pointer is retargeted from "co-evolution" in
%            the abstract to the three-node loop specification -> construction
%            -> verification; the fourth turn of the binding question appears
%            exactly once, in corollary tone, with the single clause that
%            distinguishes it from the computational turn (the certified
%            object is a deployed learning system, not an explicitly described
%            game).
%   R5'      the left fence post (equilibrium = reciprocal consistency, not a
%            unilateral robustness guarantee against nature; coincidence only
%            in zero-sum) is planted at the end of step (1); the right post
%            (consistency carries no dynamical-stability claim) is only
%            forward-referenced to 2.1 / 4.2, where it lands and works.
%   R6       step (4) is upgraded from describing the scattered literature to
%            an explicit defense of problem-oriented organization, argued from
%            the material (regret recurring across three communities).

\section{Introduction}
\label{sec:intro}

\begin{drafting}
\textbf{Introduction framing.} The introduction has to carry the reader from two
mature-but-separate fields to a single shared problem, and then to the article's
plan for attacking it from both sides.
\cmt[skeleton]{Five-step spine, revised: (1) state of game theory and AI ---
land on game-theoretic concepts as \emph{universally quantified certificates},
and plant the R5' left fence post immediately after that landing; (2) a
complementary deficit on each side, so each field naturally calls for the
other, naming the gap at the end in the vocabulary (finite data, finite
computation) that step (3) collects under ``finite access''; (3) turn to why
the gap is structurally hard, and promote the three obstructions to the
article-wide refrain: a universal certificate met by finite access, with the
regime of access (query / observational / certifier's deficit) deciding which
obstruction results (R1'); (4) the bridging effort, organized by problem, with
the organizational choice explicitly defended from the material (R6) and each
subsection tagged by its access regime; (5) close on the specification $\to$
construction $\to$ verification loop with the fourth turn as a one-sentence
corollary (R2'+R4'), then the roadmap. Guardrails: do NOT open any unit with a
``Why \dots'' question header --- it reads as self-interrogation /
self-justification. Narrate the gap \emph{before} turning to its difficulty.
Defend the organization by argument from the material, never by announcing
``this is a positioning survey.'' Keep the loop to one forward-pointing move
here; the full development belongs to Section~5.}

\cmt[step]{(1) Current state. Game theory = a language of universally
quantified certificates; AI has turned interactive in exactly the sense those
certificates quantify over. Land the move on: competence is now a strategic
question, not a scalar one. The R5' left post closes the first paragraph; the
right post is only pointed at (\ref{subsec:recap-gt} draws it,
\ref{subsec:training-alignment} uses it).}
Game theory was built to reason about situations in which an outcome depends on
how multiple agents respond to one another. Its central concepts (Nash
equilibrium, Bayesian games, incentive compatibility, rationalizability, common
knowledge, mechanism design) are all ways of disciplining this dependence. Each
quantifies over what was \emph{not} observed: whether a proposed outcome would
survive deviations no one took, whether the beliefs that could have generated
the play are consistent, whether truthful reporting would remain optimal against
every possible misreport, whether an institution would still perform once
participants understand and adapt to its rules. The logical form is the same in
each case: a universal statement ranging over counterfactual interaction. Game
theory is a language of \emph{universally quantified certificates} for
interactive systems.

One boundary of this language must be fixed before anything is built on it,
because its central certificate invites a conflation. A Nash equilibrium
certifies \emph{reciprocal} consistency: every player simultaneously best
responds to every other, so the profile is a fixed point of mutual adaptation.
Robust optimization makes a different demand, asking a single decision-maker to
perform well against an adversarial environment; the two conditions coincide
only in two-player zero-sum games, where equilibrium play reduces to minimax
robustness. Section~\ref{subsec:recap-gt} draws this fence on both of its
sides. The second side, that a consistency condition asserts nothing about
whether any learning dynamics settle on it, supplies the explanation in
Section~\ref{subsec:training-alignment} of why equilibrium-seeking training can
fail to converge.

That language has become newly urgent because artificial intelligence has turned
interactive in precisely the sense the certificates quantify over. Modern AI
systems act in environments shaped by other agents and by responses to their own
behavior. Reinforcement-learning agents learn through repeated feedback;
self-play systems reach superhuman strength in chess, Go, and poker by improving
against changing opponents~\cite{Silver2018,BrownSandholm2018}; language-model
agents negotiate, persuade, delegate, and coordinate; and pricing,
recommendation, and trading algorithms operate in environments shaped by other
algorithms and by human responses to them. Once a system acts in such a setting,
its competence must be measured strategically: how does it behave when others
adapt to it, exploit it, imitate it, or organize around it?

\cmt[step]{(2) The complementary deficit. GT specifies but cannot construct at
scale; AI constructs but cannot say what behavior should satisfy. Each lack is
the other's strength, so the two fields call for each other. Name the gap at the
END of this move, in terms (finite data, finite computation) that step (3)
will gather under ``finite access,'' so the refrain's definition is earned
rather than dropped in.}
AI and game theory meet at complementary deficits. Game theory supplies precise
strategic specifications of the kind just described, but once the number of
agents, types, or states grows large, it cannot \emph{construct} the procedures
that reach them at scale. AI supplies scalable procedures for search, function
approximation, simulation, and language-based generation, yet offers no
principled account of what their interactive behavior should \emph{satisfy}.
Game theory tends toward specification, AI toward construction. The gap between
them runs between strategic conditions quantified over counterfactual
interaction and computational procedures that must learn, approximate, or verify
those conditions from finite data and finite computation.

\cmt[step]{(3) Why hard --- and the article-wide refrain (R1'). Pivot on ``This
gap is hard for a structural reason.'' Derivation order: universal certificate
$\to$ finite access $\to$ the regime of access decides the obstruction. The
three obstructions of the previous draft (computational / epistemic /
verification) are preserved but re-derived: query access yields the
computational obstruction, observational access the epistemic one, and the
certifier's deficit of access relative to the builder yields the verification
one. On-/off-path language survives only inside the observational case. End on
the promise that the refrain is instantiated regime by regime across the six
subsections; the Part I/II bullets below redeem that promise.}
This gap is hard for a structural reason, and the reason organizes the rest of
the article. Every nontrivial strategic condition is a universally quantified
certificate: it ranges over the deviations not taken, the types not reported,
the opponents not encountered, the private values not observed, a
counterfactual space that grows combinatorially in the number of agents,
actions, and states. Any computational procedure, by contrast, reaches that
space only through what we will call \emph{finite access}: finitely many
evaluations of an explicit model, finitely many rollouts of a simulator, or
finitely many observations of play that actually occurred. The theoretical
object quantifies over more possibilities than any finite window can contain,
and which obstruction this mismatch produces depends on the regime of access.

Under \emph{query access}, the game is explicitly described, or a simulator or
differentiable model can evaluate any single counterfactual on demand. Every
individual deviation is checkable at a price, and the obstruction is
\emph{computational}: the quantified space cannot be exhausted, and assembling
finitely many local checks into the global certificate is itself intractable,
since computing a Nash equilibrium of a general finite game is PPAD-complete
and the hardness persists down to the two-player
case~\cite{DaskalakisGoldbergPapadimitriou2009,ChenDengTeng2009}. Under
\emph{observational access}, no counterfactual can be queried at any price:
only realized play is available, and the deviations not taken have left no
trace in it. The obstruction is \emph{epistemic}: the same finite record of
behavior is consistent with many distinct payoff structures, belief
hierarchies, and reasoning rules, so conduct does not uniquely reveal the
private structure that produced it. What the literature calls off-path behavior
is the case worth naming here, because it is exactly the part of the
certificate that observation never delivers.

A third obstruction cuts across both regimes, because construction and
certification are rarely performed by the same party with the same access.
Whoever must certify a system typically holds less access than whoever built
it: finite tests against an opaque policy, finite rounds of interaction with a
stronger prover, finite behavioral records from a deployed agent. The
obstruction is then one of \emph{verification}: whether a universally
quantified property can be certified from finite access at all.

This confrontation between a universal certificate and a finite window of
access is the refrain of the article. Each of the six domains surveyed below
instantiates it in a definite regime, named in each case, and closing the gap
in any regime means restructuring how a specification, its construction, and
its verification relate to one another.
\cmt[refrain]{Reverse lookup for the body sections, to keep the tags below
consistent: query regime $\to$ \ref{subsec:constructive-equilibrium},
\ref{subsec:mechanism-design}; observational regime $\to$
\ref{subsec:strategic-inference}, \ref{subsec:deployment-governance};
intermediate / verification-dominated $\to$ \ref{subsec:strategic-evaluation},
\ref{subsec:training-alignment}. Section~5 collects the refrain into the
certificate definition and taxonomy (R2') and closes the loop.}

\cmt[step]{(4) Bridging effort, organized by problem --- now defended (R6), not
merely described. Two prongs: (a) organized by community, the shared question
stays implicit seven times over; (b) a method takes its shape from its access
regime and from where it relocates the certificate, and community of origin
predicts neither --- witness the same regret quantity recurring as solution
method (3.1), training loss (3.3), and audit statistic (4.3). Then the
two-direction breakdown, Part I / Part II, which doubles as the article
structure; every bullet carries its regime tag.}
Work at this interface is abundant, but it is scattered across communities
(equilibrium computation, multi-agent reinforcement learning, empirical
game-theoretic analysis, neural mechanism design, behavioral modeling, AI
alignment, algorithmic governance), each with its own methods and benchmarks.
A survey must choose its organizing axis, and the material itself argues for
one choice. Organized by community, the literature resolves into seven internal
histories, and the question they share stays implicit in every one of them: how
can a universally quantified strategic property be represented inside a system
with finite access? Organized by problem, that question becomes the unit of
exposition, and a regularity appears that the community partition hides: a
method takes its shape from the access regime it works in and from the place to
which it relocates the certificate, and its community of origin predicts
neither. The same regret quantity recurs as a solution method in equilibrium
computation, as a training loss in neural auction design, and as an audit
statistic in algorithmic governance; an equilibrium solver and a debate
protocol, communities apart, face the same problem of certifying a universal
property from finitely many checks. We therefore organize by problem and
examine the gap from both directions.

\paragraph{Part I: AI for game theory (Section 3).}
Here AI helps make strategic specifications executable. The key observation is
that AI relocates the game-theoretic constraint into a differentiable residual,
a no-regret trajectory, an architectural guarantee, or a checkable certificate.
\begin{itemize}
  \item \textbf{Constructive equilibrium (3.1).} The Nash equilibrium is easy to
        define and hard to compute. The regime is query access: the game is
        explicit or simulatable, any deviation can be evaluated, and the
        obstruction is computational. We trace a spectrum of methods that
        relax, approximate, transform, or re-prove the equilibrium certificate,
        from differentiable residuals and no-regret self-play to
        language-model-based symbolic solvers. \emph{Given that the equilibrium
        certificate cannot be enforced globally, what detour around it does
        each method take?}
  \item \textbf{Strategic inference (3.2).} Observed behavior is consistent with
        many latent games. The regime is observational: counterfactual play
        cannot be queried, only the realized record is available, and the
        obstruction is epistemic. We examine how AI turns behavioral theories
        into trainable \emph{behavior generators} that can be constructed,
        inverted, and expanded, recovering hidden preferences, beliefs, and
        social relations. \emph{Given that fitting behavior is easy but
        identifying the mechanism is hard, what identification discipline does
        AI supply?}
  \item \textbf{Mechanism design (3.3).} Incentive compatibility requires
        truthful reporting to dominate every misreport, for every type and every
        opponent. The regime is again query access, since the learner's inner
        loop can search for profitable misreports on demand, while the
        quantified space of types and misreports remains inexhaustible. We
        trace how this counterfactual constraint moves step by step inside the
        learning system: first as an external linear program, then as a
        differentiable regret penalty~\cite{Duetting2019}, then as an
        architectural guarantee, and finally as a verified institutional
        artifact. \emph{Where should incentive compatibility live in a neural
        mechanism?}
\end{itemize}

\paragraph{Part II: Game theory for AI (Section 4).}
Here game theory identifies the relational properties that must be evaluated,
optimized, and governed once AI systems interact. The unit of analysis becomes
a policy in relation to other policies, users, protocols, and institutions.
\begin{itemize}
  \item \textbf{Strategic evaluation (4.1).} When performance depends on the
        opponent, a scalar benchmark score stops being a meaningful measure of
        competence. The evaluator occupies an intermediate regime: query access
        to the policy under evaluation, but only a finite population standing
        in for the opponents of deployment. Game-theoretic notions such as
        exploitability, population-level rankings, and Nash
        averaging~\cite{Balduzzi2018} replace fixed test sets with
        opponent-relative evaluation. \emph{Can strategic competence be judged
        without specifying against whom?}
  \item \textbf{Training and alignment (4.2).} When the alignment target cannot
        be reduced to a fixed scalar reward, for instance when it involves an
        adversary, a non-transitive preference, or a verifier weaker than the
        model~\cite{Christiano2017,Ouyang2022}, training becomes equilibrium
        selection. The verification obstruction enters the training loop
        itself: a bounded judge must certify a stronger model from finitely
        many rounds of interaction. We examine three game families:
        adversarial, preference, and interactive-proof games. \emph{When
        alignment is a relation among policies, what should training optimize?}
  \item \textbf{Deployment governance (4.3).} Individually aligned agents can
        induce undesirable system-level equilibria when they interact through
        markets, delegation chains, or pricing algorithms. The regime is
        observational in its pure form: the auditor holds behavioral records
        alone and must certify an equilibrium-level property from finite
        observation of deployed learners. No-regret learning theory and
        calibrated-regret auditing provide tools for governing strategic
        \emph{behavior} from the outside, without requiring inspection of
        source code. \emph{What system-level equilibria emerge from locally
        aligned agents, and how can they be audited?}
\end{itemize}

\cmt[step]{(5) Close on the loop (R2'+R4'), then the roadmap. The forward
pointer now names the three-node circuit specification $\to$ construction
$\to$ verification; the co-evolution argument itself still lives in Section~5.
The fourth turn of the binding question appears exactly once, in corollary
tone, carrying the one clause that keeps it distinct from the computational
turn (the certified object is a deployed learning system, no longer an
explicitly described game); the full differentiation is argued in Section~5
against the arc of \ref{subsec:recap-gt}.}
\cmt[todo]{Synchronize with conclusion.tex when it is revised under the same
mapping: its step (2) should restate the loop with the two directions assigned
to their dominant regimes (Part~I largely query, Part~II ending in pure
observation), and its closing paragraph should carry the fourth-turn sentence
with the explicit contrast to the third turn (R4'), so this corollary sentence
and that paragraph stay one argument told twice at different lengths.}
Read from both sides, these efforts close into a single circuit that
Section~\ref{sec:synthesis} develops in full: strategic specifications become
constructible; constructions, once deployed among adaptive others, generate the
demand for new specifications; and the circuit closes through verification, the
certification of a universal property from whatever access remains. One
corollary of this circuit already frames the article's end. The binding
question of game theory has turned several times, from whether an equilibrium
exists, to how an institution should be designed, to what a solution costs to
compute; the interface surveyed here turns it once more, to whether a strategic
guarantee can be constructed and checked inside a deployed learning system, an
object that no longer arrives as an explicitly described game. The remainder of
the paper proceeds as follows. Section~2 reviews the game-theoretic concepts
and AI methods the analysis requires. Section~3 studies \emph{AI for game
theory} through constructive equilibrium (3.1), strategic inference (3.2), and
neural mechanism design (3.3). Section~4 studies \emph{game theory for AI}
through strategic evaluation (4.1), training and alignment (4.2), and
deployment governance (4.3). Section~5 names the object the six domains pass
back and forth, the checkable strategic certificate, and concludes with what
its circulation implies for the design of AI systems, the governance of
multi-agent environments, and the future of game theory itself.
\end{drafting}