\section{Recap: Game Theory and Deep Learning}
\label{sec:recap}

\begin{drafting}
\textbf{Recap framing.} This section follows each field along its own
developmental path and stops where the two paths begin to meet. Game theory is
traced from the existence of equilibrium to the cost of computing it; deep
learning is traced from the function-approximation idea to the systems that now
act and reason at scale. The account is meant to be read on its own; the
concepts the rest of the article relies on appear in their native context rather
than as a checklist.
\cmt[goal]{Two changes from a pure ``preliminaries'' section: (i) tell each
discipline's story as a history with its own logic, so 2.x reads independently of
3.x/4.x; (ii) keep every forward reference (``this is what 3.1 uses'') inside
\cmt, never in the prose. The two arcs are chosen to converge: game theory ends
on \emph{computation} (algorithmic game theory), deep learning ends on
\emph{construction at scale}, and Section~\ref{sec:intro}'s gap sits exactly
between them.}
\cmt[style]{Follow introduction.tex, not ai4gt/gt4ai: narrative paragraphs with
\cmt[step] markers between movements, no Leading-question / Thesis / itemize
scaffolding. One topic sentence opens each movement; no summary-only closers.}
\end{drafting}


\subsection{Game Theory: From Existence to Computation}
\label{subsec:recap-gt}

\begin{drafting}
\cmt[arc]{Five movements, each a real turn in the field's history: (1) founding
on zero-sum + utility; (2) Nash's generalization to arbitrary games; (3)
information, dynamics, and refinement; (4) the design turn (mechanism design);
(5) the computational turn (AGT). The narrative spine is the slow shift of the
binding question from ``does an equilibrium exist'' to ``can one be
constructed.''}

\cmt[step]{(1) Founding: a mathematics of interaction, born zero-sum.}
Game theory began as the search for a principled way to act when one's payoff
depends on the choices of others. Von Neumann's minimax theorem settled the
two-player zero-sum case by showing that every such game has a value and a pair
of optimal mixed strategies, a saddle point neither player can improve
upon~\cite{Neumann1928}. The field acquired its scope when von Neumann and
Morgenstern grounded payoffs in an axiomatic theory of expected utility and
proposed game-theoretic reasoning as a foundation for economic
behavior~\cite{vNM1944}. At this stage the central solved object was the value of
a strictly competitive game.

\cmt[step]{(2) Nash: the leap to general-sum, and existence via fixed point. Land
on the certificate, because everything later relaxes or computes it.}
Nash extended the theory to games that are neither two-player nor zero-sum. A
Nash equilibrium is a strategy profile in which each player already best-responds
to the others, so no one can gain by deviating alone; using a fixed-point
argument, Nash proved that every finite game has at least one such equilibrium in
mixed strategies~\cite{Nash1951}. Existence thereby became automatic, and the
content of the concept settled onto the condition it certifies: a global
guarantee that no unilateral deviation pays. This no-profitable-deviation
certificate, and the question of how to reach it, organizes most of what
follows.
\cmt[anchor]{Define exploitability / the Nash gap here in one clause so 3.1 and
4.1 can reuse it without re-introducing: the worst-case gain available to a
best-responding deviator, zero exactly at equilibrium.}

\cmt[step]{(3) Information, dynamics, refinement: the post-Nash elaboration that
won the 1994 + 2005 Nobels. Keep it as one connected movement, not a list.}
The decades after Nash enriched the equilibrium concept along three fronts. The
extensive form gave games a temporal structure of moves and information sets, and
Kuhn's theorem related behavior strategies to mixed strategies under perfect
recall~\cite{Kuhn1953}. Because Nash equilibrium permits incredible threats off
the equilibrium path, Selten introduced refinements, subgame and trembling-hand
perfection, that demand rationality at every contingency~\cite{Selten1975}.
Harsanyi brought private information inside the theory by modeling uncertainty
about payoffs as uncertainty about a player's type drawn from a common prior,
turning games of incomplete information into ordinary Bayesian
games~\cite{Harsanyi1968}. A parallel epistemic program asked what players must
believe about one another: Aumann formalized common knowledge and showed that
correlation through a shared signal yields a broader and often more natural
notion, the correlated equilibrium~\cite{Aumann1976,Aumann1974}. Where perfect
rationality strained against evidence, weaker behavioral concepts appeared:
quantal response equilibrium replaces exact optimization with a smooth, noisy
best response~\cite{McKelveyPalfrey1995}, and rationalizability keeps only the
strategies that survive iterated removal of choices that are never a best
response~\cite{Bernheim1984,Pearce1984}.
\cmt[forward]{Quiet hooks for later, not stated in prose: correlated / coarse
correlated equilibrium = the certificates no-regret dynamics actually reach
(3.1, 4.3); QRE = the differentiable response map inverted in 3.2. Mention them
in their own historical right here; let 3.x/4.x cite back.}

\cmt[step]{(4) The design turn: invert the question. Predict play -> design rules.
This is the ``for game theory'' engineering face and the natural home of IC.}
A second line inverted the analytic question. Mechanism design takes the desired
outcome as given and asks how to set the rules of a game so that self-interested
play produces it. The ambition met early limits:
Arrow showed that no voting rule can aggregate preferences while satisfying a
short list of reasonable axioms~\cite{Arrow1951}, and Gibbard and Satterthwaite
showed that any non-dictatorial choice rule over unrestricted preferences is
manipulable~\cite{Gibbard1973,Satterthwaite1975}. Within these limits a
constructive theory grew. The revelation principle lets a designer restrict
attention to direct mechanisms in which truthful reporting is an equilibrium, so
the design problem reduces to imposing incentive compatibility and individual
rationality. The Vickrey--Clarke--Groves mechanisms achieve efficient outcomes in
dominant strategies by charging each agent its externality on the
others~\cite{Vickrey1961,Clarke1971,Groves1973}, and Myerson characterized the
revenue-maximizing auction for a single item~\cite{Myerson1981}. Incentive
compatibility is the counterfactual constraint at the center of this theory:
truthful reporting must remain optimal against every misreport an agent could
have made.
\cmt[anchor]{This IC definition + the VCG / affine-maximizer family are exactly
what 3.3 relocates step by step; introduce them as economics here so 3.3 can lean
on them without redefining.}

\cmt[step]{(5) The computational turn: from existence to construction. This is
where the survey picks up. Land the complexity partition hard.}
As the games and mechanisms of interest grew large, the binding question moved
from whether a solution exists to whether one can be computed, and algorithmic
game theory formed around that question~\cite{NisanRoughgardenTardosVazirani2007}.
The answer depends sharply on the setting. A two-player zero-sum equilibrium
coincides with the minimax value and is computable by linear programming, hence
efficient. Computing a Nash equilibrium of a general game is far harder: the
problem is PPAD-complete, a class capturing fixed-point problems whose solutions
are guaranteed to exist but appear intractable to find, and the hardness persists
even for two players~\cite{Papadimitriou1994,DaskalakisGoldbergPapadimitriou2009,ChenDengTeng2009}.
Classical exact solvers such as Lemke--Howson run in worst-case exponential
time~\cite{LemkeHowson1964}. The correlated solution concepts are by contrast
benign: a correlated or coarse correlated equilibrium can be computed in
polynomial time, and is reached by simple decentralized learning,
since the empirical play of agents who each minimize regret converges to this
set~\cite{HartMasColell2000,BlumMansour2007}. How much such self-interested
learning costs society is itself bounded, by the price of anarchy in games with
appropriate smoothness~\cite{KoutsoupiasPapadimitriou1999,Roughgarden2015,SyrgkanisTardos2013}.
Mechanism design carries an analogous computational burden, since constructing
optimal incentive-compatible mechanisms is itself a hard optimization over the
space of type-contingent rules~\cite{ConitzerSandholm2002}.
\cmt[bridge]{This paragraph is the hand-off to the whole paper, but say it in one
later sentence, not here: zero-sum tractability and PPAD-hardness are why 3.1's
solvers target two-player zero-sum and retreat to (C)CE elsewhere; no-regret ->
correlated equilibrium is the hinge to 4.3; AGT is the field this survey extends.}

The shift this section traces, from establishing that an equilibrium exists to
asking at what cost it can be built, is precisely the opening that learning-based
methods step into.
\end{drafting}


\subsection{Deep Learning: From Function Approximation to Systems That Act}
\label{subsec:recap-dl}

\begin{drafting}
\cmt[arc]{Five movements paralleling 2.1's logic: (1) the network as a trainable
function approximator; (2) scale turns capacity into capability; (3) architecture
as inductive bias; (4) generative modeling; (5) decision-making and language. End
where the substrate becomes capable of \emph{constructing} the objects 2.1
priced.}

\cmt[step]{(1) Foundations: connectionism -> backprop -> universal approximation.
Frame the network as a differentiable parameterized map. This is the technical
primitive the whole survey rests on.}
Deep learning rests on a single idea: represent an unknown mapping as a composition
of simple adjustable units and fit it to data. The early connectionist model, the
perceptron, learned a linear separator by adjusting weights from
examples~\cite{Rosenblatt1958}, but Minsky and Papert showed that a single layer
cannot represent even elementary nonlinear functions, and interest
lapsed~\cite{MinskyPapert1969}. The obstruction fell once networks with hidden
layers were trained by backpropagation, which computes the gradient of a loss
with respect to every weight by the chain rule~\cite{Rumelhart1986}, and once it
was proved that a network with a single hidden layer of sufficient width can
approximate any continuous function on a compact
domain~\cite{Cybenko1989,Hornik1991}. A neural network is thus a parameterized,
differentiable map, and training is the minimization of an empirical loss by
gradient descent. This is the primitive on which everything later is built.

\cmt[step]{(2) Scale: capacity -> capability. The 2012 inflection. Representation
learning replaces hand-engineered features.}
Approximation capacity alone did not make networks useful; scale did. Larger
datasets and faster hardware turned depth into a practical advantage, a shift
marked by convolutional networks for images~\cite{LeCun1998} and by the
large-margin improvement of a deep network on the ImageNet benchmark, which moved
the field decisively toward learned representations over hand-engineered
features~\cite{Krizhevsky2012}. Stochastic gradient descent with adaptive step
sizes made the optimization of very large models routine~\cite{Kingma2015Adam},
and the consolidation of these advances established deep networks as the default
approximator across domains~\cite{LeCunBengioHinton2015,GoodfellowBengioCourville2016}.

\cmt[step]{(3) Architecture as inductive bias. The lesson: match structure to
data; the architecture encodes a prior that holds for every input.}
The next lesson was that the form of the network matters as much as its size,
because a well-chosen architecture builds the structure of the problem into the
model. Convolution encodes spatial locality and translation invariance; attention
and the transformer let every element of a set or sequence condition on every
other, removing the assumption of a fixed input order~\cite{Vaswani2017}; and
equivariant designs make a network respect symmetries of its input by
construction. An architectural choice acts as a prior that holds for all inputs,
without being learned from data or enforced by the loss.

\cmt[step]{(4) Generative modeling: learn to sample, not just to predict. Flag
GAN as itself a game in one clause; full treatment is 4.2.}
A parallel line of work learned the data distribution itself, so that new
samples can be drawn from it. Variational autoencoders optimize a
likelihood bound through an encoder--decoder pair~\cite{KingmaWelling2014};
generative adversarial networks pit a generator against a discriminator and
reach their target when the discriminator can no longer tell real data from
generated~\cite{Goodfellow2014}; and diffusion models generate by learning to
reverse a gradual noising process~\cite{Ho2020DDPM}. The adversarial formulation
is notable for casting generation as a two-player game, a framing the alignment
literature would later adopt directly.

\cmt[step]{(5) Decision-making and language: from prediction to action, then to
symbolic/linguistic competence. Close on LLMs as both actors and engines.}
Two further developments extended learned approximation from prediction to action
and language. Reinforcement learning treats behavior as a policy in a Markov
decision process and improves it from reward, either through estimated values or
through the policy gradient, which differentiates expected return with respect to
the policy's parameters~\cite{Williams1992,SuttonPolicyGradient2000,SuttonBarto2018};
combined with deep function approximation, this produced agents that learn control
from raw perception~\cite{Mnih2015} and, through self-play, reach superhuman
strength in games of pure skill~\cite{Silver2018}. A related strand recovers the
objective itself from observed behavior, through inverse reinforcement learning
and its adversarial imitation variants~\cite{NgRussell2000,HoErmon2016}. Most
recently, transformers trained at scale to predict the next token in text have
become large language models that acquire in-context learning and the generation
of code and structured arguments~\cite{Vaswani2017,BrownGPT3_2020}, extending the
same differentiable-approximation machinery to symbolic and linguistic tasks.

By the end of this path the network can search, generate, act, and write
programs, and that capability is what the next section turns toward the
construction of game-theoretic objects.
\end{drafting}